---
title: "Week 4 – Statistics: Distributions & First Model"
author: "Group ID: ODS-TEAM-14"
date: "`r format(Sys.Date())`"
format:
  html:
    toc: false
    toc-depth: 2
editor: 
  markdown: 
    wrap: 72
---

## Overview {.unnumbered}

::: callout-note
### What this file is for

This week is hands-on with **statistics**. You will: - Measure spread
(**SD, IQR**) for two numerical variables - Visualise **distribution
shapes** (histograms + Normal overlays) - Fit a **simple linear
regression** (e.g., `grade ~ study_hours`) - Inspect **residuals** to
check model fit - Record brief **group reflections**

Work with your **cleaned Week 2 dataset** (no impossible or extreme
values).
:::

------------------------------------------------------------------------

# Dataset & Setup

Use your cleaned dataset prepared in Week 2. Mind the file path.

```{r}
library(tidyverse)

# Load your cleaned dataset
#install.packages("yfR")
#install.packages("scales")
library(yfR)

#The below path is for PC
data <- read.csv("~/ods-2025/ods-group-project-ods-team-14/data/processed/processed_data.csv")

#The below path is for github and it can be replaced with the PC path to access the dataset.

# Quick structure check
glimpse(data)
```

```{r}
summary(data)
```

**Confirm (1–2 lines):** that the dataset loaded correctly
(rows/columns) and that `grade` and `study_hours` are numeric.

Ans --> Dataset loaded successfully with 5736 rows and 10 columns both Daily_Return and Range are numeric , so can be used for statistical analysis and modeling.
------------------------------------------------------------------------

# Spread & Variability

Compute summaries, SD, and IQRs for at least two two main variables from
your dataset (e.g., `grade` and `study_hours`).

```{r}
# Summaries
# descriptive statistics
sd_iqr <- data.frame(
Variable = c("Daily_Return", "Range"),
SD = c(sd(data$Daily_Return), sd(data$Range)),
IQR = c(IQR(data$Daily_Return), IQR(data$Range))
)
sd_iqr
```

**Comment (2–3 lines):** Which variable varies more? Are the ranges
reasonable? What does the **IQR** say about the middle 50%?

Ans --> Range varies more than Daily_Return
as it has higher SD and IQR. Daily return stays close to zero while range fluctuates widely.
------------------------------------------------------------------------

# Distribution Shapes

Plot histograms and compare with a Normal curve for the same variables.

```{r}
# Histogram and normal curve for variable 1
library(ggplot2)

# Histogram and normal curve for Daily_Return
hist(data$Daily_Return,breaks = 40,prob = TRUE,
     main = "Distribution of Daily Returns",
     xlab = "Daily Return",
     col = "skyblue",
     border = "white")
curve(dnorm(x,mean = mean(data$Daily_Return, na.rm = TRUE),
            sd = sd(data$Daily_Return, na.rm = TRUE)),
      add = TRUE, col = "red", lwd = 2)
 
# Histogram and normal curve for Range
hist(data$Range,breaks = 40,prob = TRUE,
     main = "Distribution of Range",
     xlab = "Range",
     col = "lightgreen",
     border = "white")
 
curve(dnorm(x,mean = mean(data$Range, na.rm = TRUE),
            sd = sd(data$Range, na.rm = TRUE)),
      add = TRUE, col = "red", lwd = 2)
```

**Comment (2–3 lines):** Is each variable roughly Normal, skewed, or
heavy-tailed? Does the red curve fit well? Any gaps or outliers?

Ans --> The range is right-skewed and heavy-tailed as well as the daily_return is closely to normal.Overall the distribution is slightly heavy-tailed.
------------------------------------------------------------------------

# First Model (Simple Linear Regression)

Fit a simple linear model (use appropriate variables and reason your
choice).

```{r}
# create a model usinf lm and provide summary
model <- lm(data$Daily_Return ~ data$Range)
```

```{r}
summary(model)
```

**Interpretation (3–4 lines):** - Intercept (β₀) = \_\_\_ means … -
Slope (β₁) = \_\_\_ means … - (Optional) Is the slope statistically
significant (p-value)? - (Optional) How much variation does the model
explain (R²)?

Ans --> Intercept (β₀) = 0.00203 means when range = 0, then daily_return is approximately = 0.20%
Slope (β₁) = -7.46 means the daily_return descreases showing large intraday functions.

::: callout-note
### What is a model?

A **model** is a simplified, mathematical explanation of how things
relate or behave.\
It is **expressed mathematically** (equation/algorithm), **summarises**
relationships, **simplifies** what matters, **generalises (within
limits)**, and **must be interpretable**.

> A good model is clear, honest, and useful — not a perfect copy of
> reality, but a way to think better about it.
:::

------------------------------------------------------------------------

# Model Fit (Residuals)

Residuals are **errors** = observed − predicted. They should look like
random noise around 0.

```{r}
# Plot the residuals including abline
plot(model$residuals, 
     main = "Model Residuals", 
     ylab = "Residuals")
abline(h = 0, col = "red")
```

**Comment (2–3 lines):** Are residuals roughly centred around 0 and
randomly scattered? Any patterns or extreme points? What does this
suggest about model fit?

Ans --> The residual is linearly fit and is centered around 0 and it is randomly scattered above and below the red line. It has no random strong patterns which makes the linear relationship between Daily_return and range to be reasonable.
------------------------------------------------------------------------

# Group Reflection

Keep answers short and specific (1–2 sentences each).

| Question | Reflection |
|:-----------------------------------|:-----------------------------------|
| **Distributions:** What did shapes reveal about variation or consistency?

|  | --> The residual is randomly scattered and roughly centered around zero and a consistent variance is across the fitted values.

| **Model:** What does your regression line suggest about the relationship?

|  | --> The regression line depicts that there is a weak negative correlation between Range and Daily_Return i.e. the higher the Range the lower is the Daily_Return.

| **Learning:** What was the main thing you learned about your dataset this week? |  |

--> We learned how to build and interpret a simple linear regression model, and how to understand basic statistical terms like SD (standard deviation) and IQR (interquartile range), which helped to observe stock prices change over time.
------------------------------------------------------------------------

# Expected Outputs

Your file should include (at least): - Two distribution plots - One
scatterplot with fitted regression line - One residual plot - Regression
summary with brief interpretation - Completed reflection table

------------------------------------------------------------------------

# You are free to explore further along modeling (e.g., trying different variable combinations, or trying multiple regression, or different models available in R libraries).

# You may also explore along p-values and R².
